---
title: "SPEAR + Lasso: all tasks"
author: "Jeremy Gygi (Steven Kleinstein Lab + Leying Guan Lab)"
output: html_notebook
---

Packages

```{r}
# For SPEAR: (https://bitbucket.org/kleinstein/SPEAR)
library(SPEAR)
library(MultiAssayExperiment)
# For lasso:
library(glmnet)
# For plotting:
library(ggplot2)
library(UpSetR)
# For imputation:
library(MOFA2)
```

Data

```{r}
train.data <- readRDS("../master_processed_training_data.RDS")
data <- readRDS("../master_harmonized_training_data.RDS")

test.data <- readRDS("../master_processed_prediction_data.RDS")
data.te <- readRDS("../master_processed_prediction_data.RDS")
```

Preprocessing: Convert tibbles into computable matrices (remove specimen_id column and make rownames)

```{r}
# Perform initial loading:
# SPEAR needs matrices
ab_titers <- as.matrix(t(train.data$abtiter_wide$normalized_data))
olink <- as.matrix(t(train.data$plasma_cytokine_concentrations$normalized_data))
cytof <- as.matrix(t(train.data$pbmc_cell_frequency$normalized_data))
gex <- as.matrix(t(train.data$pbmc_gene_expression$batchCorrected_data))

# Test data:
ab_titers.te <- as.matrix(t(test.data$abtiter$processed_similar_to_training))
olink.te <- as.matrix(t(test.data$plasma_cytokine_concentrations$processed_similar_to_training))
cytof.te <- as.matrix(t(test.data$pbmc_cell_frequency$processed_similar_to_training))
gex.te <- as.matrix(t(test.data$pbmc_gene_expression$processed_similar_to_training))
```

Preprocessing: Check distributions

ab_titers:

```{r}
# Ab Titers:
ab_titers_preprocessed <- apply(ab_titers, 2, log2)
ab_titers_preprocessed_sc <- scale(ab_titers_preprocessed)
ab_titers_preprocessed.te <- apply(ab_titers.te, 2, log2)
ab_titers_preprocessed_sc.te <- scale(ab_titers_preprocessed.te)

# OLINK:
# NOTE: Removing IL4 (less than 0...)
olink_preprocessed <- olink[,colnames(olink)!="P05112"]
olink_preprocessed <- apply(olink_preprocessed, 2, log2)
olink_preprocessed_sc <- scale(olink_preprocessed)
# NOTE: Removing IL4 (less than 0...)
olink_preprocessed.te <- olink.te[,colnames(olink.te)!="P05112"]
olink_preprocessed.te <- apply(olink_preprocessed.te, 2, log2)
olink_preprocessed_sc.te <- scale(olink_preprocessed.te)

# CyTOF:
cytof_preprocessed <- apply(cytof, 2, sqrt)
cytof_preprocessed_sc <- scale(cytof_preprocessed)
cytof_preprocessed.te <- apply(cytof.te, 2, sqrt)
cytof_preprocessed_sc.te <- scale(cytof_preprocessed.te)

# Gene Expression:
half.min = min(gex[gex != 0])/2
gex_preprocessed <- apply(gex + half.min, 2, log2)
gex_preprocessed_sc <- scale(gex_preprocessed)
half.min.te = min(gex.te[gex.te != 0])/2
gex_preprocessed.te <- apply(gex.te + half.min.te, 2, log2)
gex_preprocessed_sc.te <- scale(gex_preprocessed.te)
```

Check dimensions and number of samples:

```{r}
print("Train")
dim(ab_titers_preprocessed_sc)
dim(olink_preprocessed_sc)
dim(cytof_preprocessed_sc)
dim(gex_preprocessed_sc)
print("Test")
dim(ab_titers_preprocessed_sc.te)
dim(olink_preprocessed_sc.te)
dim(cytof_preprocessed_sc.te)
dim(gex_preprocessed_sc.te)
```

analyte overlap:

```{r}
# ab_titers:
tr.analytes <- colnames(ab_titers_preprocessed_sc)
te.analytes <- colnames(ab_titers_preprocessed_sc.te)
which(!tr.analytes %in% te.analytes)
tr.analytes <- colnames(olink_preprocessed_sc)
te.analytes <- colnames(olink_preprocessed_sc.te)
which(!tr.analytes %in% te.analytes)
tr.analytes <- colnames(cytof_preprocessed_sc)
te.analytes <- colnames(cytof_preprocessed_sc.te)
which(!tr.analytes %in% te.analytes)
tr.analytes <- colnames(gex_preprocessed_sc)
te.analytes <- colnames(gex_preprocessed_sc.te)
which(!tr.analytes %in% te.analytes)
```

Note that CyTOF from test data has two extra features:

```{r}
# Confirm overlap:
tr.analytes <- colnames(cytof_preprocessed_sc)
te.analytes <- colnames(cytof_preprocessed_sc.te)
colnames(cytof_preprocessed_sc.te)[which(!te.analytes %in% tr.analytes)]
# Remove them from test:
cytof_preprocessed_sc.te <- cytof_preprocessed_sc.te[,-c(which(!te.analytes %in% tr.analytes))]
# Check again:
tr.analytes <- colnames(cytof_preprocessed_sc)
te.analytes <- colnames(cytof_preprocessed_sc.te)
which(!te.analytes %in% tr.analytes)
```

specimen_id overlap:

```{r}
specimen_ids_list <- list(gex = rownames(gex_preprocessed), ab_titers = rownames(ab_titers_preprocessed), olink = rownames(olink_preprocessed), cytof = rownames(cytof_preprocessed))
specimen_ids_list.te <- list(gex = rownames(gex_preprocessed.te), ab_titers = rownames(ab_titers_preprocessed.te), olink = rownames(olink_preprocessed.te), cytof = rownames(cytof_preprocessed.te))

UpSetR::upset(data = UpSetR::fromList(specimen_ids_list))
UpSetR::upset(data = UpSetR::fromList(specimen_ids_list.te))
```

Get overlapping subjects:

```{r}
comb_ids <- dplyr::intersect(
              dplyr::intersect(
                dplyr::intersect(specimen_ids_list$gex, specimen_ids_list$ab_titers), 
                  specimen_ids_list$olink), 
              specimen_ids_list$cytof)
comb_ids.te <- dplyr::intersect(
              dplyr::intersect(
                dplyr::intersect(specimen_ids_list.te$gex, specimen_ids_list.te$ab_titers), 
                  specimen_ids_list.te$olink), 
              specimen_ids_list.te$cytof)



# Which are baseline?
comb_visits <- sapply(comb_ids, function(s_id){
  return(data$subject_specimen$visit[which(data$subject_specimen$specimen_id == s_id)])
})
visit1_comb_ids <- comb_ids[which(comb_visits == 1)]
comb_visits.te <- sapply(comb_ids.te, function(s_id){
  return(data.te$subject_specimen$planned_day_relative_to_boost[which(data.te$subject_specimen$specimen_id == s_id)])
})
visit1_comb_ids.te <- comb_ids.te[which(comb_visits.te == 0)]

# Participants:
visit1_comb_participants <- sapply(visit1_comb_ids, function(s_id){
  return(data$subject_specimen$subject_id[which(data$subject_specimen$specimen_id == s_id)])
})
visit1_comb_participants.te <- sapply(visit1_comb_ids.te, function(s_id){
  return(data.te$subject_specimen$subject_id[which(data.te$subject_specimen$specimen_id == s_id)])
})

# Some test participants are missing a baseline...
all.te <- unique(data.te$subject_specimen$subject_id)
full.te <- unname(visit1_comb_participants.te)
missing.te <- all.te[!all.te %in% full.te]
print(length(missing.te))
print(missing.te)

# Do these 2 individuals have full timepoints prior?
dplyr::filter(data.te$subject_specimen, subject_id %in% missing.te)

# Unfortunately some don't... need to impute...
```

# Quick MOFA imputation:

```{r}
datasets_combined = list(
  ab_titers = ab_titers_preprocessed_sc.te,
  olink = olink_preprocessed_sc.te,
  cytof = cytof_preprocessed_sc.te,
  gex = gex_preprocessed_sc.te
)

generate_full_matrices = function(datasets_combined){
  event_ids <- sort(unique(unlist(sapply(datasets_combined, function(i){return(rownames(i))}))))
  new_datasets_combined <- list()
  for(i in 1:length(datasets_combined)){
    print(i)
    d <- datasets_combined[[i]]
    m <- matrix(NA, nrow = length(event_ids), ncol = ncol(d))
    rownames(m) <- event_ids
    colnames(m) <- colnames(d)
    m <- as.data.frame(m)
    m[rownames(d),] <- d
    new_datasets_combined[[i]] <- m
  }
  names(new_datasets_combined) <- names(datasets_combined)
  return(new_datasets_combined)
}

datasets_full_combined <- generate_full_matrices(datasets_combined)



# Run MOFA+:
x.mofa <- list()
for(d in 1:length(datasets_full_combined)){
  x.mofa[[d]] = t(datasets_full_combined[[d]])
}
names(x.mofa) = names(datasets_full_combined)
MOFAobject <- create_mofa(x.mofa)
# MOFA+ specific parameters:
data_opts <- get_default_data_options(MOFAobject)
model_opts <- get_default_model_options(MOFAobject)
train_opts <- get_default_training_options(MOFAobject)
model_opts$num_factors = 63
train_opts$convergence_mode = "fast"
train_opts$startELBO = 2
train_opts$maxiter = 1000
train_opts$stochastic = FALSE
train_opts$seed = 42
train_opts$drop_factor_threshold = -1
MOFAobject <- prepare_mofa(
  object = MOFAobject,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)
MOFAobject.trained <- run_mofa(MOFAobject,use_basilisk = TRUE, save_data = F)

imputedMOFA <- MOFA2::predict(MOFAobject.trained, factors = "all", add_intercept = T)
# Replace any NA in data with imputedMOFA:
tmp = datasets_full_combined
names(tmp) <- names(datasets_full_combined)
for(i in 1:length(tmp)){
  non_observed <- is.na(tmp[[i]])
  tmp[[i]][non_observed] = t(imputedMOFA[[i]][[1]])[non_observed]
}
datasets_full_combined <- tmp
lapply(datasets_full_combined, dim)
```

Save/Load Imputed Data

```{r}
#saveRDS(datasets_full_combined, "prediction_imputed_data.rds")
datasets_full_combined <- readRDS("prediction_imputed_data.rds")
```

Preparing Test Data:

```{r}
# Check all 63 samples are full:
specimen_ids_list.te <- list(gex = rownames(datasets_full_combined$gex), ab_titers = rownames(datasets_full_combined$ab_titers), olink = rownames(datasets_full_combined$olink), cytof = rownames(datasets_full_combined$cytof))
UpSetR::upset(data = UpSetR::fromList(specimen_ids_list.te))

# Get baseline for predictions:
comb_ids.te <- dplyr::intersect(
              dplyr::intersect(
                dplyr::intersect(specimen_ids_list.te$gex, specimen_ids_list.te$ab_titers), 
                  specimen_ids_list.te$olink), 
              specimen_ids_list.te$cytof)
length(comb_ids.te) # 63


# Which are baseline?
comb_visits.te <- sapply(comb_ids.te, function(s_id){
  return(data.te$subject_specimen$planned_day_relative_to_boost[which(data.te$subject_specimen$specimen_id == s_id)])
})
visit1_comb_ids.te <- comb_ids.te[which(comb_visits.te == 0)]
length(visit1_comb_ids.te) # 21

# Participants:
visit1_comb_participants.te <- sapply(visit1_comb_ids.te, function(s_id){
  return(data.te$subject_specimen$subject_id[which(data.te$subject_specimen$specimen_id == s_id)])
})
```

Obtain response values for SPEAR training:

```{r}
# Train:
# Getting responses for tasks:
visit1_comb_participants <- sapply(visit1_comb_ids, function(s_id){
  return(data$subject_specimen$subject_id[which(data$subject_specimen$specimen_id == s_id)])
})

# Task 1: day 14 IgG_PT:
task1.Y <- sapply(visit1_comb_participants, function(s_id){
  specimen_id_day14 <- data$subject_specimen$specimen_id[which(data$subject_specimen$subject_id == s_id & data$subject_specimen$visit == 5)]
  return(ab_titers_preprocessed[as.character(specimen_id_day14),"IgG_PT"])
})
# Task 1: day 14 IgG_PT FC (vs. day 0):
task1.Y.FC <- sapply(visit1_comb_participants, function(s_id){
  specimen_id_day14 <- data$subject_specimen$specimen_id[which(data$subject_specimen$subject_id == s_id & data$subject_specimen$visit == 5)]
  specimen_id_day0 <- data$subject_specimen$specimen_id[which(data$subject_specimen$subject_id == s_id & data$subject_specimen$visit == 1)]
  return(ab_titers_preprocessed[as.character(specimen_id_day14),"IgG_PT"] - ab_titers_preprocessed[as.character(specimen_id_day0),"IgG_PT"])
})

# Task 2: day 1 Monocytes:
task2.Y <- sapply(visit1_comb_participants, function(s_id){
  specimen_id_day1 <- data$subject_specimen$specimen_id[which(data$subject_specimen$subject_id == s_id & data$subject_specimen$visit == 2)]
  return(cytof_preprocessed[as.character(specimen_id_day1),"Monocytes"])
})
# Task 2: day 1 Monocytes: FC (vs. day 0):
task2.Y.FC <- sapply(visit1_comb_participants, function(s_id){
  specimen_id_day1 <- data$subject_specimen$specimen_id[which(data$subject_specimen$subject_id == s_id & data$subject_specimen$visit == 2)]
  specimen_id_day0 <- data$subject_specimen$specimen_id[which(data$subject_specimen$subject_id == s_id & data$subject_specimen$visit == 1)]
  return(sqrt(cytof[as.character(specimen_id_day1),"Monocytes"]/cytof[as.character(specimen_id_day0),"Monocytes"]))
})

# Task 3: day 3 CCL3:
task3.Y <- sapply(visit1_comb_participants, function(s_id){
  specimen_id_day3 <- data$subject_specimen$specimen_id[which(data$subject_specimen$subject_id == s_id & data$subject_specimen$visit == 3)]
  return(gex_preprocessed[as.character(specimen_id_day3),"ENSG00000277632.1"])
})
# Task 3: day 3 CCL3: FC (vs. day 0):
task3.Y.FC <- sapply(visit1_comb_participants, function(s_id){
  specimen_id_day3 <- data$subject_specimen$specimen_id[which(data$subject_specimen$subject_id == s_id & data$subject_specimen$visit == 3)]
  specimen_id_day0 <- data$subject_specimen$specimen_id[which(data$subject_specimen$subject_id == s_id & data$subject_specimen$visit == 1)]
  return(gex_preprocessed[as.character(specimen_id_day3),"ENSG00000277632.1"] - gex_preprocessed[as.character(specimen_id_day0),"ENSG00000277632.1"])
})

# Scale tasks:
task1.Y <- scale(task1.Y)
task1.Y.FC <- scale(task1.Y.FC)
task2.Y <- scale(task2.Y)
task2.Y.FC <- scale(task2.Y.FC)
task3.Y <- scale(task3.Y)
task3.Y.FC <- scale(task3.Y.FC)
```

Molding data into SPEAR-ready format:

```{r}
assay_ab_titers <- ab_titers_preprocessed_sc[visit1_comb_ids,]
assay_olink <- olink_preprocessed_sc[visit1_comb_ids,]
assay_cytof <- cytof_preprocessed_sc[visit1_comb_ids,]
assay_gex <- gex_preprocessed_sc[visit1_comb_ids,]

assay_ab_titers.te <- datasets_full_combined$ab_titers[visit1_comb_ids.te,]
assay_olink.te <- datasets_full_combined$olink[visit1_comb_ids.te,]
assay_cytof.te <- datasets_full_combined$cytof[visit1_comb_ids.te,]
assay_gex.te <- datasets_full_combined$gex[visit1_comb_ids.te,]

dim(assay_ab_titers)
dim(assay_olink)
dim(assay_cytof)
dim(assay_gex)

dim(assay_ab_titers.te)
dim(assay_olink.te)
dim(assay_cytof.te)
dim(assay_gex.te)
```

Model Training:

```{r}
# Function to get clinical values:
get.clinical.values <- function(taskname, clinical.values = NULL){
  res <- list()
  for(val in clinical.values){
    if(val == "baseline"){
      if(taskname == "task1.1" | taskname == "1_1"){
        res[["baseline"]] <- assay_ab_titers[,"IgG_PT"]
      } else if(taskname == "task1.2" | taskname == "1_2") {
        res[["baseline"]] <- assay_ab_titers[,"IgG_PT"]
      } else if(taskname == "task2.1" | taskname == "2_1") {
        res[["baseline"]] <- assay_cytof[,"Monocytes"]
      } else if(taskname == "task2.2" | taskname == "2_2") {
        res[["baseline"]] <- assay_cytof[,"Monocytes"]
      } else if(taskname == "task3.1" | taskname == "3_1") {
        res[["baseline"]] <- assay_gex[,"ENSG00000277632.1"]
      } else if(taskname == "task3.2" | taskname == "3_2") {
        res[["baseline"]] <- assay_gex[,"ENSG00000277632.1"]
      }
    } else if(val == "age"){
      tmp <- data$subject_specimen[rownames(assay_ab_titers),]
      approx.age <- sapply(1:nrow(tmp), function(idx){
        return(as.integer(difftime(as.Date("2022-01-01"), tmp$year_of_birth[idx], units = "days")/365))
      })
      res[["approx_age"]] <- approx.age
    } else if(val == "sex"){
      tmp <- data$subject_specimen[rownames(assay_ab_titers),]
      res[["biological_sex"]] <- ifelse(tmp$biological_sex == "Male", 1, 0)
    } else if(val == "vac"){
      tmp <- data$subject_specimen[rownames(assay_ab_titers),]
      res[["infancy_vac"]] <- ifelse(tmp$infancy_vac == "wP", 1, 0)
    }
  }
  return(do.call("cbind", res))
}

# Response
Y.list <- list(
  task1.1 = task1.Y,
  task1.2 = task1.Y.FC,
  task2.1 = task2.Y,
  task2.2 = task2.Y.FC,
  task3.1 = task3.Y,
  task3.2 = task3.Y.FC
)

# How many SPEAR factors:
k = 10

# Names for files to save models:
modelnames <- c(paste0("spear_task_1_1_k", k, ".rds"),
                paste0("spear_task_1_2_k", k, ".rds"),
                paste0("spear_task_2_1_k", k, ".rds"),
                paste0("spear_task_2_2_k", k, ".rds"),
                paste0("spear_task_3_1_k", k, ".rds"),
                paste0("spear_task_3_2_k", k, ".rds"))

# Run SPEAR and Ensemble?
do.spear <- TRUE
do.ensemble <- TRUE

for(i in 1:6){
  # list of assays (rows = analytes, cols = samples!)
  assays.tr <- list(olink = t(assay_olink),
                 cytof = t(assay_cytof),
                 gex = t(assay_gex),
                 ab_titers = t(assay_ab_titers))
  Y.tr <- Y.list[[i]]
  spear.modelname <- modelnames[i]
  # add response in colData:
  df.tr <- data.frame(Response = as.vector(Y.tr),
                   row.names = rownames(assay_ab_titers))

  # MAE:
  mae.tr <- MultiAssayExperiment::MultiAssayExperiment(experiments = assays.tr,
                                                    colData = df.tr)
  if(do.spear){
      SPEARobj <- SPEAR::new.spear(data = mae.tr,
                                         response = "Response",
                                         num.factors = k,
                                         num.folds = 2,
                                         print.out = 5)
    
      # perform leave-one-out cv training:
      SPEARobj$train.spear()
      saveRDS(SPEARobj, file = spear.modelname)
  }
  
  if(do.ensemble){
    SPEARobj <- readRDS(spear.modelname)
    SPEARobj$set.weights(method = "min")
    fs <- SPEARobj$get.factor.scores()
    Xcomb <- cbind(fs, get.clinical.values(taskname = names(Y.list)[i], clinical.values = c("baseline", "age", "sex", "vac")))
    colnames(Xcomb) <- c(colnames(fs), "Baseline", "Approx_age", "Biological_sex", "Infancy_vac")
    lasso_fit = glmnet::cv.glmnet(x = Xcomb, y = Y.tr, foldid = 1:46)
    #saveRDS(lasso_fit, file = gsub("spear", "ensemble", spear.modelname))
  }
}
```

Generate Predictions:

```{r}
get.clinical.values.te <- function(taskname, clinical.values = NULL){
  res <- list()
  for(val in clinical.values){
    if(val == "baseline"){
      if(taskname == "task1.1" | taskname == "1_1"){
        res[["baseline"]] <- assay_ab_titers.te[,"IgG_PT"]
      } else if(taskname == "task1.2" | taskname == "1_2") {
        res[["baseline"]] <- assay_ab_titers.te[,"IgG_PT"]
      } else if(taskname == "task2.1" | taskname == "2_1") {
        res[["baseline"]] <- assay_cytof.te[,"Monocytes"]
      } else if(taskname == "task2.2" | taskname == "2_2") {
        res[["baseline"]] <- assay_cytof.te[,"Monocytes"]
      } else if(taskname == "task3.1" | taskname == "3_1") {
        res[["baseline"]] <- assay_gex.te[,"ENSG00000277632.1"]
      } else if(taskname == "task3.2" | taskname == "3_2") {
        res[["baseline"]] <- assay_gex.te[,"ENSG00000277632.1"]
      }
    } else if(val == "age"){
      tmp <- data.te$subject_specimen[which(data.te$subject_specimen$specimen_id %in% as.integer(rownames(assay_ab_titers.te))),]
      approx.age <- sapply(1:nrow(tmp), function(idx){
        return(as.integer(difftime(as.Date("2022-01-01"), tmp$year_of_birth[idx], units = "days")/365))
      })
      res[["approx_age"]] <- approx.age
    } else if(val == "sex"){
      tmp <- data.te$subject_specimen[which(data.te$subject_specimen$specimen_id %in% as.integer(rownames(assay_ab_titers.te))),]
      res[["biological_sex"]] <- ifelse(tmp$biological_sex == "Male", 1, 0)
    } else if(val == "vac"){
      tmp <- data.te$subject_specimen[which(data.te$subject_specimen$specimen_id %in% as.integer(rownames(assay_ab_titers.te))),]
      res[["infancy_vac"]] <- ifelse(tmp$infancy_vac == "wP", 1, 0)
    }
  }
  return(do.call("cbind", res))
}

assays.te <- list(olink = t(assay_olink.te),
                 cytof = t(assay_cytof.te),
                 gex = t(assay_gex.te),
                 ab_titers = t(assay_ab_titers.te))
df.te <- data.frame(Response = rep(NA, length = nrow(assay_ab_titers.te)),
                    row.names = rownames(assay_ab_titers.te))
participant_ids_te <- unname(visit1_comb_participants.te[rownames(assay_ab_titers.te)])

# MAE:
mae.te <- MultiAssayExperiment::MultiAssayExperiment(experiments = assays.te,
                                                    colData = df.te)

preds <- list()
for(i in 1:6){
  SPEARmodel <- readRDS(modelnames[i])
  SPEARmodel$set.weights(method = "min")
  SPEARmodel$add.data(data = mae.te, name = "test")
  fs.te <- SPEARmodel$get.factor.scores(data = "test")
  Xcomb.te <- cbind(fs.te, get.clinical.values.te(taskname = names(Y.list)[i], clinical.values = c("baseline", "age", "sex", "vac")))
  colnames(Xcomb.te) <- c(colnames(fs.te), "Baseline", "Approx_age", "Biological_sex", "Infancy_vac")
  lasso_fit = readRDS(gsub("spear", "ensemble", modelnames[i]))
  tmp = stats::predict(lasso_fit, Xcomb.te, s = "lambda.min")
  rownames(tmp) <- participant_ids_te
  colnames(tmp) <- names(Y.list)[i]
  preds[[i]] <- tmp
}

preds.df <- do.call("cbind", preds)
#saveRDS(preds.df, "010924_SPEAR_ensemble_predictions_final.rds")
```

Finally, write table for predictions:

```{r}
preds.df <- readRDS("010924_SPEAR_ensemble_predictions_final.rds")
preds.df.ranked <- apply(preds.df, 2, function(col){return(rank(-col))})
#write.csv(preds.df.ranked, "SPEAR_ensemble_predictions.csv")
```
